\section{Conclusions and future work}\label{sec:conclusions}

In this paper we described our ongoing work in conceiving the next generation of electrical and software architecture of vehicles of the near future. The work is performed in the context of two Swedish projects led by Volvo Cars. In particular, in this work we focus on how to architect a car as a constituent system of a future transportation system. The outcome of this research will define a specific viewpoint of the Volvo Cars architecture framework~\cite{JSA2017}.

As future work we aim at defining precise model kinds that will enable a proper description of the system of systems viewpoint of cars of the near future.
%In this paper we describe the main building blocks of cars that want to play a role in a system of systems. 

Moreover, as discussed in the paper, %shown by the concerns emerging from the concepts spanning the entire system of systems, 
engineering efforts are %not only needed on the different system levels, but also on 
needed also at the level of the system of systems as a whole. When constituent systems are developed by different organizations, which might have different owners, managers, and goals, other aspects need to be investigated. In general, there should be an engineering effort at the level of the whole SoS from the definition of the goals of the SoS, to the development, management, and maintenance of the constituent systems. Constituent systems should be able to join the SoS, to properly interact and interoperate with the other constituent systems, and to perform all the actions that are required to guarantee the achievement of the SoS goals, even at the cost of temporarily sacrificing the constituents' goals. Finally, the collaboration and interoperability of constituent systems in a SoS might generate potentially dangerous emergent behaviours that should be properly controlled and avoided.   
%First of all we should understand who should be the ``owner" of the SoS, as well as who is defining the goals and requirements of the SoS, and according to which process. 

%being developed by a different organization, a separate set of questions arise with respect to the coordination of engineering efforts: 
%
%\begin{itemize}
%\item How shall system of system goals and requirements on these be defined? 
%\item How shall creation of requirements on the system of systems as a whole be coordinated? 
%\item What processes are needed to handle development requests between separate systems? 
%\item How shall development of interoperability specifications be coordinated? 
%\item How shall the SoS services be supervised, managed and maintained over the different life cycles of the constituents?
%\end{itemize}
%
%Within the project we will try to give an answer to these questions.