\section{Engineer an SoS: challenges and opportunities}

Connected vehicles will benefit from Intelligent Transport Systems (ITS), Smart
Cities, and Internet of Things (IoT). 
Future transportation systems will be a heterogeneous mix of items with varying connectivity and interoperability. A mix of new technologies and legacy systems will co-exist to realize a variety of scenarios involving not only connected vehicles but also smart roads, pedestrians, cyclists, and so on. In other words, future transportation systems are expected to be system of systems, where each constituent system can act as a standalone system, but where the cooperation among the constituent systems will enable new emerging and promising scenarios. 


Connected vehicles  will combine data from inside vehicle
with external data coming from the environment (other vehicles, the road, signs,
and the cloud). In such a scenario, different applications will be possible: smart
traffic control, better platooning coordination, and enhanced safety in general.
In today's vehicles, the driver plays a fundamental role. Besides controlling
the vehicle, she/he is also expected to recover the vehicle from failures. With fully
autonomy (SAE level 5), the driver is no longer in the loop. The vehicle, besides
controlling the basic driving maneuvers, should handle all possible situations during
driving. As the driver is not monitoring the vehicle and is not a fallback option
anymore as in the lower levels of automation, it is expected that nothing goes wrong.
For the higher levels of automation (i.e., levels 3 - 5), additional players enter the
ecosystem: companies that focus on communication aspects will play a big role in
this ecosystem, as communication will be an important basis for highly automated
driving~\cite{Knauss2017challenges}. Communication allows the vehicles to ensure that they have redundant
information sources (i.e., several sources of information) to ensure safety, reliability,
and quality of autonomous vehicles. As can be seen in this example, data plays an
essential role in autonomous vehicles. Hence, companies that focus on data will be
additional players in the ecosystem~\cite{Knauss2017challenges}. When it comes to OEMs for highly automated
driving, we notice that companies need to shift to being software companies
instead of developing software just as a side product, as already mentioned above.


As anticipated in Section~\ref{sec:sota}, SoSE is an emerging discipline of the last years that addresses the development,
operation and maintenance of SoSs. Even though the community around SoSE is ac trive, SoSE is a
young discipline, and general principles and theoretical foundations remain to be discovered. The two
biggest challenges are related to: 

\begin{itemize}
\item Uncertainty about constituents - SoS need to manage unavoidable imprecise
and uncertain information about constituent systems; 
\item Independence of constituents - constituent systems
are independent entities for what concern operation, management, and evolution. 
\end{itemize}

Each constituent has its own
objectives, possibly conflicting with objectives of the other constituents or of the SoS. An SoS has a potentially
limited authority that it exercises on the constituent systems to drive them towards its objectives. 

An open problem of SoSs is how to provide justification of the reliance on emergent properties~\cite{Fitzgerald2014}. Emergent
properties are the result of synergistic collaboration between constituent systems, and are those properties
that cannot be expressed at the level of a single constituent system, but require observations of phenomena at the
SoS boundary~\cite{Fitzgerald2014}. There are two different types of emergence: weak emergence [6], when an emergent property
can be reduced to its constituents, and strong emergence, when an emergent property cannot be traced to any
direct cause (a classic example is the consciousness emerging as a property of the brain~\cite{Fitzgerald2014}). Like~\cite{EmergenceAndRefinement,Fitzgerald2014}  we
confine our attention to weak emergence, and hereafter, we use the term emergence to refer to weak emergence.
Emergence can be either anticipated, meaning that it is defined at design-time, or unanticipated, meaning
that it is not purposely or consciously designed-in or surprising to the developers and users of the SoS. Consequences
of the unanticipated emergent behavior may be viewed as negative/harmful, positive/beneficial, or
neutral/unimportant by stakeholders of the SoS~\cite{DDUSA2012}. Existing approaches to design and verify dependable systems\footnote{Dependability is a generic concept including attributes like reliability, availability, safety, integrity, maintainability, etc.~\cite{dependability}.} 
work pretty well with closed and unchanging systems. These techniques become inadequate for assessing
and justifying SoS reliance on emergent properties since SoSs are composed of autonomous constituents
whose behaviour can neither be predicted nor controlled~\cite{Strigini12}. Investigation is needed to provide ways for characterizing, quantifying and measuring a system
behavior in the presence of perturbations~\cite{Strigini12}.

One of the most frequently used techniques to explore emergence is simulation~\cite{DeWolf2005}. The work in~\cite{Zambonelli04} advocates that different engineering techniques are required at different scales,
including software engineering and formal methods. Self-adaptive systems~\cite{de2013software,Author:2009tt,Salehie2009} are systems that autonomously
decide how to adapt at runtime to meet environment (context) and user changes and threats. Uncertainty about
constituents and independence of constituents make most of the techniques proposed for self-adaptive system
inapplicable for SoSs. Providing guarantees that an SoS will achieve a global goal in predictable and dependable
ways, through the collaboration of autonomous and independent constituent systems with different (and
potentially conflicting) local goals remains an open problem~\cite{Delgado2004,Leucker2009293,ZHANG20185}. There is the need of evidence on which
to base reliance on emergent properties and on SoS-level behaviours.

 



\patrizio{autonomous cars are becoming part of an autonomous system of system. An autonopmous car needs to costantly communicate with other cars and the infrastracture, cia sensors and the cloud. unprecedented complexity will lead to unprecedented cooperation \url{https://www.ibm.com/blogs/internet-of-things/iot-design-for-system-of-systems}. Enginenrign the SoS: coopration bewetween companies, + standards, etc.\\
Anche: ``An interconnected self-driving car system-of-systems"\\
Anche: Toyota - Intelligent transport systems: \url{www.toyota-global.com/innovation/smart_mobility_society/intelligent_transport_systems}\\
Anche the EU project SOCRATES2.0 (System of Coordinated Roadside and Automotive Services for Traffic Efficiency and Safety), partnership of BMW, HERE, and Copenhagen
}

?	Should the SoS be Virtual, Collaborative, Acknowledged, or Directed, or a mix of them?
?	Should the collaboration be established for a limited time in response to or to solve specific situations, like hazards in connected safety scenarios?
?	Should the collaboration be extemporary and spontaneous?
?	Who is engineering the SoS, one specific OEM, a group of OEMs, road authorities, etc.?
?	Who is the owner of the SoS?
?	Who is the manager of the SoS?


The
research significance is many-fold since SoS Engineering is an emerging discipline of the last years and is still in
his infancy. SoS might be used in several safety-critical application domains, e.g., intelligent transport systems,
integrated air defense networks, applications in healthcare, emergency response, space exploration, homeland
security, etc. Consequently, SoS engineering needs to be grounded on a rigorous conceptual framework based
on mathematical principles, so to enable: (i) precise and unambiguous definition of SoS properties; (ii) runtime
verification so to control, avoid, or block negative/harmful unanticipated emergent behaviours, (iii) the development
of tools for managing the SoS, (iv) industry to develop products and services with guaranteed properties
while excluding unwanted behaviors. The understanding of the impact of proposed or potential changes will
enable SoS system engineers to either intervene in order to preclude problems or develop strategies to mitigate
the impact on the SoS. Moreover, the runtime verification methods will enable the definition of self-adaptation
methods for the SoS, according to the type of SoS, i.e. Virtual, Collaborative, Acknowledged, or Directed, and
in general to the extent of ?controllability? of constituents.
In our discussions with several Swedish companies (between others Volvo Car Group, Volvo Trucks, Ericsson,
Axis communications, Jeppesen - Boing) all of them claimed that one of the business challenge for industry
in the near future is to build and provide new functionalities and/or services in a short time scale by easily integrating
the system with other systems and to continuously maintain and evolve the system under guarantees of
quality of service.



